# Draft: Results

## Overview of the Study Population

The analysis of the tornado data from the NOAA Storm Events Database spans from 1950 through 20205. After combining individual storm segments into complete tornado events, a total of 31,345 unique tornado events were identified across the United States. Of these events, 761 started within the defined 3-mile coastal zone. These events were categorized by their trajectory to determine the potential for transporting sharks. The population breakdown is as follows in @tbl-shark-results:

```{r}
#| echo: false
#| warning: false
#| label: tbl-shark-results
library(gt)
library(tidyverse)

shark_results_tbl <- tibble(
  path_movement = c("Moving Coastal", "Moving Coastal", "Moving Inland", "Moving Inland", "Other Path", "Other Path"),
  shark_lifting_potential = c("High Potential", "Low Potential", "High Potential", "Low Potential", "High Potential", "Low Potential"),
  event_count = c(2, 60, 2, 71, 3799, 27411),
  percentage = c(0.0323, 0.968, 0.0274, 0.973, 0.122, 0.878)
)

shark_results_tbl %>%
  gt(groupname_col = "path_movement") %>%
  tab_header(
    title = "Tornado Path Trajectory and Shark-Lifting Potential",
    subtitle = "1950-2025"
  ) %>%
  tab_caption("A table caption") %>%
  cols_label(
    shark_lifting_potential = "Shark-Lifting Potential",
    event_count = "Event Count",
    percentage = "Proportion of Trajectory"
  ) %>%
  fmt_percent(
    columns = percentage,
    decimals = 1
  ) %>%
  fmt_number(
    columns = event_count,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  tab_options(
    table.width = pct(80),
    heading.title.font.size = px(20)
  )
```

Initial findings indicate that while "High Potential" and "Moving Inland" events are a minority of total tornado occurrences, with only **2 total** tornadoes meeting the criteria for "High Potential" and "Moving Inland". Thus, these unique events are certainly are not impossible and present a specific kind of threat that current risk models might be missing.

TODO add figure caption

## Path Movement & Intensity Analysis

To evaluate the relationship between tornado path trajectory and intensity, the data was categorized by both path movement and its "Shark-Lifting" potential (with F3/F4/F5 tornadoes marked as "High Potential"). As seen in Figure \_ below, there is a significant difference in tornadoes that remained completely inland ("Other Path") compared to tornadoes that either originated or ended on the coasts. While "Moving Inland" trajectories are the primary focus when considering the risk of sharknadoes, on **2.7%** of these tornadoes moving from the coast to the inland reached high-potential intensity. In contrast, "Other Path" tornadoes exhibited a high-potential rate of **12.2%**.

TODO add figure caption

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(dplyr)

final_glued_events <- readRDS("final_glued_events.rds")

plot_data <- final_glued_events %>%
  count(path_movement, shark_lifting_potential) %>%
  group_by(path_movement) %>%
  mutate(percent = n / sum(n)) # Calculate relative proportion for better comparison

ggplot(plot_data, aes(x = path_movement, y = percent, fill = shark_lifting_potential)) +
  geom_bar(stat = "identity", position = "fill", color = "white") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("High Potential" = "#b2182b", "Low Potential" = "#4393c3")) +
  theme_minimal() +
  labs(
    title = "Relative Tornado Intensity by Path Trajectory",
    subtitle = "Comparing the likelihood of High Potential (F3/F4/F5) events by path trajectory",
    x = "Tornado Path Category",
    y = "Percentage of Events",
    fill = "Intensity Category"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 10)
  )
```

This suggests that while the "Moving Inland" path is geographically unique, the prime atmospheric conditions needed for F3/F4/F5 tornadoes are far more prevalent once the storm has moved more inland over land. For insurance risk modeling, this indicates that while coastal-origin tornadoes are *capable* of lifting and moving sharks, they are statistically much less likely to reach the extreme wind speeds of their inland counterparts.

## Regional Risk Distribution

To identify likely areas where we could find sharknadoes forming, the data was segmented by coastal region (Atlantic, Gulf, and Pacific). The analysis reveals a geographic imbalance in risk.

The **Gulf Coast** and **Atlantic Coast** account for the vast majority of "Moving Inland" events, with each coast responsible for 36 "Moving Inland" tornadoes. In, contrast the **Pacific Coast** only recorded **1 total** "Moving Inland" tornado which was identified as "Low Potential" intensity. In all, though, the majority of the "Moving Inland" tornadoes produced by any of the 3 coasts were identified as "Low Potential" intensity.

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(dplyr)
library(scales)

regional_plot_data <- final_glued_events %>%
  filter(path_movement == "Moving Inland") %>%
  group_by(Start_coast, shark_lifting_potential) %>%
  summarize(count = n(), .groups = "drop")

ggplot(regional_plot_data, aes(x = reorder(Start_coast, -count, sum), y = count, fill = shark_lifting_potential)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = c("High Potential" = "#b2182b", "Low Potential" = "#4393c3")) +
  theme_minimal() +
  labs(
    title = "Regional 'Moving Inland' Activity",
    subtitle = "Frequency of coastal-origin tornadoes by region and lifting potential",
    x = "Coastal Region",
    y = "Total Number of Events",
    fill = "Potential Category"
  ) +
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )
```
This confirms that sharknado potential is a regional phenomenon, rather than a national one, with the Atlantic and Gulf Coasts requiring significantly different actuarial considerations than the West Coast. 

## Economic Impact
The final phase of this analysis evaluated the economic impact of tornadoes based on their path trajectory. Across the 75-year study period, the total property damage recorded for all events summed to \$3,334,748,030 with an average of \$106,389 per tornado.

Again, when categorizing tornadoes by path movement, "Moving Inland" events resulted in an average property loss of \$342,153 per storm. While these events are less frequent than inland-only "Other Path" tornadoes, their tendency to originate in high-value coastal zones contributes to a distinct loss profile.

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(dplyr)
library(scales)

damage_plot_data <- final_glued_events %>%
  group_by(path_movement) %>%
  summarize(total_loss = sum(Total_Damage, na.rm = TRUE))

ggplot(damage_plot_data, aes(x = path_movement, y = total_loss, fill = path_movement)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  scale_fill_manual(values = c("Moving Inland" = "#1b9e77", 
                               "Moving Coastal" = "#d95f02", 
                               "Other Path" = "#7570b3")) +
  theme_minimal() +
  labs(
    title = "Cumulative Economic Impact by Trajectory",
    subtitle = "Total property damage (1950-2025) in Millions of USD",
    x = "Tornado Path Category",
    y = "Total Property Damage (Millions)",
    fill = "Path Category"
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))
```

Interestingly, the data indicates that while 'High Potential' (F4/F5) storms cause the most catastrophic individual losses, the sheer volume of 'Low Potential' storms moving from coast to land creates a cumulative financial burden that could require specific actuarial attention.

TODO add more charts to break down financials between categories

::: callout-note

I love the way you've written this.

:::
