# Results

## Overview of the Study Population

The analysis of the tornado data from the NOAA Storm Events Database spans from 1950 through 2025. After combining individual storm segments into complete tornado events, a total of 31,345 unique tornado events were identified across the United States. Of these events, 761 started within the defined 3-mile coastal zone, and only 73 of those ultimately moved from the coast to the inland. These events were categorized by their trajectory to determine the potential for transporting sharks. The population breakdown is as follows in @tbl-population-breakdown:

```{r}
#| label: tbl-population-breakdown
#| echo: false
#| warning: false
library(gt)
library(tidyverse)
#| echo: false
#| warning: false
#| label: tbl-population-breakdown
library(gt)
library(tidyverse)

final_glued_events <- readRDS("final_glued_events.rds")

# generate the summary data
shark_results_summary <- final_glued_events %>%
  group_by(path_movement, shark_lifting_potential) %>%
  summarise(event_count = n(), .groups = "drop") %>%
  group_by(path_movement) %>%
  mutate(percentage = event_count / sum(event_count)) %>%
  ungroup()

# table
shark_results_summary %>%
  gt(groupname_col = "path_movement") %>%
  tab_header(
    title = "Tornado Path Trajectory and Shark-Lifting Potential",
    subtitle = "1950-2025"
  ) %>%
  tab_caption("Population breakdown: categorized all tornadoes by path trajectory and calculated counts and proportions of high vs. low shark-lifting potential within each path trajectory group.") %>%
  cols_label(
    shark_lifting_potential = "Shark-Lifting Potential",
    event_count = "Event Count",
    percentage = "Proportion of Trajectory"
  ) %>%
  fmt_percent(
    columns = percentage,
    decimals = 1
  ) %>%
  fmt_number(
    columns = event_count,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  tab_options(
    table.width = pct(80),
    heading.title.font.size = px(20)
  )
```

Initial findings indicate that "High Potential" and "Moving Inland" events are a minority of total tornado occurrences, with only **2 total** tornadoes meeting the criteria for "High Potential" and "Moving Inland". While these unique events are quite rare, they are certainly not impossible and present a specific kind of threat that current risk models might be missing.

Furthermore, the analysis identified an empirical frequency of **0.0064%** for sharknado-like ("Moving Inland", high shark-lifting potential) storms. To determine the precision of this estimate, a 95% confidence interval was calculated. The resulting interval, ranging from 0.0017% to 0.023%, indicates that while the true probability is extremely low, the upper bound remains nearly four times higher than the point estimate. 

## Path Movement & Intensity Analysis

To further evaluate the relationship between tornado path trajectory and intensity, we can visualize the metrics from the previous @tbl-population-breakdown. In @fig-trajectory-intensity below, we can see a noticeable difference in tornadoes that remained completely inland ("Other Path") compared to tornadoes that either originated or ended on the coasts. While "Moving Inland" trajectories are the primary focus when considering the risk of sharknadoes, only **2.7%** of these tornadoes moving from the coast to the inland reached high-potential intensity. In contrast, "Other Path" tornadoes exhibited a much larger high-potential rate of **12.2%**.

```{r}
#| echo: false
#| warning: false
#| label: fig-trajectory-intensity
#| fig-cap: "Categorized all tornadoes by path trajectory and visualized the percentages of high vs. low shark-lifting potential within each path trajectory group."
library(ggplot2)
library(dplyr)

final_glued_events <- readRDS("final_glued_events.rds")

plot_data <- final_glued_events %>%
  count(path_movement, shark_lifting_potential) %>%
  group_by(path_movement) %>%
  mutate(percent = n / sum(n))

ggplot(plot_data, aes(x = path_movement, y = percent, fill = shark_lifting_potential)) +
  geom_bar(stat = "identity", position = "fill", color = "white") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("High Potential" = "#b2182b", "Low Potential" = "#4393c3")) +
  theme_minimal() +
  labs(
    title = "Relative Tornado Intensity by Path Trajectory",
    subtitle = "Comparing the likelihood of High Potential (F3/F4/F5) events by path trajectory",
    x = "Tornado Path Category",
    y = "Percentage of Events",
    fill = "Intensity Category"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 10)
  )
```

This suggests that while the "Moving Inland" path is geographically unique, the prime atmospheric conditions needed for F3/F4/F5 tornadoes are far more prevalent once the storm has moved more inland over land. For insurance risk modeling, this indicates that while coastal-origin tornadoes are *capable* of transporting sharks, they are statistically much less likely to reach the extreme wind speeds of their inland counterparts.

## Regional Risk Distribution

To identify likely areas where we could find sharknadoes forming, the data was segmented by coastal region (Atlantic, Gulf, and Pacific). The analysis in @fig-moving-inland-regions reveals a geographic imbalance in risk.

The **Gulf Coast** and **Atlantic Coast** account for the vast majority of "Moving Inland" events, with each coast responsible for 36 "Moving Inland" tornadoes. In contrast, the **Pacific Coast** only recorded **1 total** "Moving Inland" tornado, which was identified as "Low Potential" intensity. The **Atlantic Coast** was the only coastal region that had recorded a "Moving Inland" tornado with **high shark-lifting potential** (2 total occurrences recorded). Overall, though, the majority of the "Moving Inland" tornadoes produced by any of the 3 coasts were identified as "Low Potential" intensity.

```{r}
#| echo: false
#| warning: false
#| label: fig-moving-inland-regions
#| fig-cap: "Categorized only 'Moving Inland' trajectory tornadoes by coastal region and visualized the percentages of high vs. low shark-lifting potential within each coastal region group."
library(ggplot2)
library(dplyr)
library(scales)

regional_plot_data <- final_glued_events %>%
  filter(path_movement == "Moving Inland") %>%
  group_by(Start_coast, shark_lifting_potential) %>%
  summarize(count = n(), .groups = "drop")

ggplot(regional_plot_data, aes(x = reorder(Start_coast, -count, sum), y = count, fill = shark_lifting_potential)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = c("High Potential" = "#b2182b", "Low Potential" = "#4393c3")) +
  theme_minimal() +
  labs(
    title = "Regional 'Moving Inland' Activity",
    subtitle = "Frequency of coastal-origin tornadoes by region and lifting potential",
    x = "Coastal Region",
    y = "Total Number of Events",
    fill = "Potential Category"
  ) +
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )
```
This confirms that sharknado potential is a regional phenomenon, rather than a national one, with the Atlantic and Gulf Coasts requiring significantly different actuarial considerations than the West Coast. 

## Economic Impact
The final phase of this analysis evaluated the economic impact of tornadoes based on their path trajectory. Across the 75-year study period, the total property damage recorded for all events summed to **\$3.33B** with an average of \$106,389 per tornado. The summary of economic impact by tornado trajectory and shark-lifting potential is in @tbl-trajectory-damage below: 

```{r}
#| echo: false
#| warning: false
#| label: tbl-trajectory-damage

library(gt)
library(dplyr)
library(scales)

table_data <- final_glued_events %>%
  group_by(path_movement, shark_lifting_potential) %>%
  summarize(
    event_count = n(),
    total_loss = sum(Total_Damage, na.rm = TRUE),
    avg_loss = mean(Total_Damage, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(path_movement, desc(shark_lifting_potential))

table_data %>%
  gt(groupname_col = "path_movement") %>%
  tab_header(
    title = "Economic Impact Analysis by Tornado Trajectory",
    subtitle = "Property damage and frequency (1950-2025)"
  ) %>%
  tab_caption("Categorized all tornadoes by path trajectory and calculated total and average property damage inflicted by high vs. low shark-lifting potential tornadoes within each path trajectory group.") %>%
  cols_label(
    shark_lifting_potential = "Shark-Lifting Potential",
    event_count = "Total Events",
    total_loss = "Cumulative Damage",
    avg_loss = "Avg. Damage per Event"
  ) %>%
  fmt_currency(
    columns = c(total_loss, avg_loss),
    currency = "USD",
    suffixing = TRUE 
  ) %>%
  fmt_number(
    columns = event_count,
    decimals = 0
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  tab_options(
    table.width = pct(90),
    heading.title.font.size = px(20),
    row.striping.include_table_body = FALSE,
    row_group.background.color = "white",
    table.background.color = "white"
  )
```

As visualized in @fig-trajectory-damage-total below, tornadoes following "Other Path" trajectories have contributed to **`r round(((682.77+2580)/3334) * 100, 2)`%** of property damage in the last 75 years, which makes sense since they are the most commonly occurring tornadoes. In contrast, tornadoes that originated on the coast and moved inland ("Moving Inland") have contributed to only **`r round(((24.79+0.183)/3334) * 100, 2)`%** of all recorded tornado property damage, with only **$183K** in damages resulting from high shark-lifting potential tornadoes under the "Moving Inland" trajectory category. 


```{r}
#| echo: false
#| warning: false
#| label: fig-trajectory-damage-total
#| fig-cap: "Categorized all tornadoes by trajectory and visualized the total property damage inflicted by high vs. low shark-lifting potential tornadoes within each path trajectory group."

library(ggplot2)
library(dplyr)
library(scales)

damage_plot_data <- final_glued_events %>%
  group_by(path_movement, shark_lifting_potential) %>%
  summarize(total_loss = sum(Total_Damage, na.rm = TRUE), .groups = "drop")


ggplot(damage_plot_data, aes(x = path_movement, y = total_loss, fill = shark_lifting_potential)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) + 
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  scale_fill_manual(values = c("High Potential" = "#b2182b", "Low Potential" = "#4393c3")) +
  theme_minimal() +
  labs(
    title = "Cumulative Economic Impact by Trajectory and Intensity",
    subtitle = "Total property damage (1950-2025) grouped by Path Trajectory and Shark-Lifting Potential",
    x = "Tornado Path Category",
    y = "Total Property Damage (Millions USD)",
    fill = "Lifting Potential"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

Based on the visual analysis provided in @fig-trajectory-damage-avg, a hasty conclusion would be that tornadoes originating inland and moving to the coast ("Moving Coastal") result in the most property damage on average. Based on the calculations, this is technically true. However, it's important to note that for "Moving Coastal" and "Moving Inland" tornadoes classified as high shark-lifting potential, the population consists only of **2 total recorded events** for both groups. 


```{r}
#| echo: false
#| warning: false
#| label: fig-trajectory-damage-avg
#| fig-cap: "Categorized all tornadoes by trajectory and visualized the average property damage inflicted by high vs. low shark-lifting potential tornadoes within each path trajectory group."

library(ggplot2)
library(dplyr)
library(scales)

# Calculate the mean instead of the sum
damage_plot_data_avg <- final_glued_events %>%
  group_by(path_movement, shark_lifting_potential) %>%
  summarize(avg_loss = mean(Total_Damage, na.rm = TRUE), .groups = "drop")

ggplot(damage_plot_data_avg, aes(x = path_movement, y = avg_loss, fill = shark_lifting_potential)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) + 
  # Using Millions (M) suffix; if values are smaller, you can remove 'scale' and 'suffix'
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  scale_fill_manual(values = c("High Potential" = "#b2182b", "Low Potential" = "#4393c3")) +
  theme_minimal() +
  labs(
    title = "Average Economic Impact by Trajectory and Intensity",
    subtitle = "Mean property damage (1950-2025) grouped by Path Trajectory and Shark-Lifting Potential",
    x = "Tornado Path Category",
    y = "Average Property Damage (Millions USD)",
    fill = "Lifting Potential"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

Ultimately, when looking at the tornadoes with sharknado characteristics ("Moving Inland" with high shark-lifting potential), only **$183K** in damages has been inflicted in the last 75 years with an average damage of **\$91.5K per storm**. From a risk management perspective, this indicates that while high-intensity coastal-to-inland events are physically possible, their historical financial impact is negligible compared to standard inland tornadic activity. For an insurance agency, these figures suggest that "Sharknado-style" events represent an extremely rare risk rather than a primary cause of annual losses from catastrophes.

:::{.callout-note}
A confidence interval could not be calculated for the average damage of "Moving Inland", high shark-lifting potential tornadoes due to the extreme scarcity of data $(n=2)$. In these instances, the point estimate of $91.5K per storm serves as a historical record rather than a predictive mean.
:::

## Additional Analysis

To further analyze the historical trends of tornado weather intensity, the (Enhanced) Fujita scale classifications for every recorded tornado from 1950-2025 were plotted on the line graph shown in @fig-historical-intensity-trends below. In the figure, there is a clear spike in F0 and F1 tornadoes after 1995. Higher intensity tornadoes remained relatively constant throughout the time period. Possible explanations for this feature in the data can be found in @sec-conclusion-freq.

```{r}
#| label: fig-historical-intensity-trends
#| echo: false
#| warning: false
#| fig-cap: All tornadoes categorized by (Enhanced) Fujita category and counts are plotted from 1950-2025

library(tidyverse)

# 1. Prepare Data with Year Filter
intensity_trends_filtered <- final_glued_events %>%
  mutate(
    # Combine F and EF scales by extracting just the number
    scale_num = str_extract(TOR_F_SCALE_MAX, "\\d"),
    intensity_label = paste0("Category ", scale_num)
  ) %>%
  filter(!is.na(scale_num)) %>%
  mutate(intensity_label = factor(intensity_label, 
                                  levels = paste0("Category ", 0:5))) %>%
  count(YEAR, intensity_label)

# 2. Create the Clean Line Graph
ggplot(intensity_trends_filtered, aes(x = YEAR, y = n, color = intensity_label)) +
  # Standard lines only (no smoothing)
  geom_line(linewidth = 1) + 
  scale_color_viridis_d(option = "plasma", end = 0.9) +
  labs(
    title = "Tornado Intensity Trends (1950 - 2025)",
    subtitle = "Aggregated F and EF Scale counts by category",
    x = "Year",
    y = "Annual Count",
    color = "Intensity Category"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(1950, 2025, 5))

```
